---
title: "UFO Sightings Visualized by RYAN COST"
output: ioslides_presentation
---

<style type="text/css">body, p, div, h1, h2, h3, h4, h5 
{color:black;font-family:Avenir Pro 85 Heavy;}
slides>slide.title-slide hgroup 
h1 {color:#165a2d;}
h2 {color:#165a2d;}
body{color:#0e3a1d;}
p{font-size: 12px;}
div{color:#0e3a1d;}
</style> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Data sourcing
## https://www.kaggle.com/NUFORC/ufo-sightings

# Slide Plan:
#   1 Title page
#   2 Why this data?
#   3 Data fixing 
#       - &#44 ( , ) did not load in
#       - changing the dates for easier wrangling
#   4 First plot - Barchart ( Sighting count by State) (ggplot2)
#   5 second plot - 3d PLotly scatter (plottly) (3 Variables)
#               - x: year
#               - y: duration
#               - z: distance from middle of US (use lat/long calc)
#   6 third plot - map of US with sightings (ggplot2)
#   7 four plot - Line (Sightings over time in 3 different states) (plotly)
#   8 Stats on sighting over time (t test)
#   9 Conclusions based on the data
#   10 COnclusion / final remarks
#           

dataC = read.csv("complete.csv", sep = ",", header = TRUE)
```

```{r, include = FALSE}
library(testthat)
library(digest)
library(stringr)
library(dplyr)
library(tibble)
library(knitr)
library(tidyr)
library(ggplot2)
library(plotly)
library(geosphere)
```

## UFO Sightings.. Are We Alone?

- I found a dataset with over **89000 cases** of "Alien" contacts, all in the United States.I thought it would be intriguing to plot some of this large pool of different experiences and see how alien sightings trend across the States.
- Also, This data is large, so part of the challenge will be select relevant parts and using that to make observations via graphs/plots
- link to data: <https://www.kaggle.com/NUFORC/ufo-sightings>.
- The United States is not monlithic so a dataset that covers all states can be interesting to see how different cultures/surroundings effect observations of UFOs.
- This is also a hot topic because of the new releases from the government on declassifying UFO sightings!

## Why This Data?

- Different types of column fields - easy to make interesting inferences
- Lots of entries and each is a real entry by a person. The Data has comments which are neat to read, but don't work good to make plots
- Aliens are so interesting!


## Are There Aliens in my State??

```{r, include = FALSE}
dataC <- data.frame(dataC)

us_sightings <- filter(dataC, dataC$country == 'us')
us_sightings <- data.frame(us_sightings)

us_sightings <- us_sightings %>% group_by(state) %>% mutate(stateSightings = n())

us_sightings$year = str_sub(us_sightings$datetime,-10,-6)
us_sightings <- us_sightings %>% group_by(state, year) %>% mutate(currentSightings = n())

p<-ggplot(data=us_sightings, aes(x=state, y=stateSightings, colour = state)) +
  geom_bar(stat="unique", width = .8) +
  scale_x_discrete(guide = guide_axis(n.dodge=3))

p <- p  + guides(colour=guide_legend(nrow=4))
p <- p + theme(legend.position = "bottom") + ylab("Number of Sightings")
p <- p + ggtitle("Sightings by State") + theme(plot.title = element_text(size=22)) + theme(legend.title = element_text(size=5))


```
- California has the most sightings, in front of Washington and Florida
```{r, echo = FALSE}
p

```

## Data Wrangling / Fixing for the Bar graphs
- I used dplyr library to group and count the total sightings per state, and add it as a column in a single line 
```{r, echo = TRUE, eval = FALSE}
us_sightings <- us_sightings %>% group_by(state) 
%>% mutate(stateSightings = n())
```
- I also changed how the legend looked to make the bar graph look nicer and have more space:
```{r, echo = TRUE, eval = FALSE}
p <- p  + guides(colour=guide_legend(nrow=4))
p <- p + theme(legend.position = "bottom") 
        + ylab("Number of Sightings")
```

## How Far and How Long?

<p>This compares year to duration as well as how far the sighting is from "the center of the US".<br> This can give insight into how sightings changed over time and where sightings are located.</p><br>
```{r, include = FALSE}

#               - x: year
#               - y: duration
#               - z: distance from middle of US (use lat/long calc)


us_sightings3D <- filter(dataC, dataC$country == 'us', dataC$durationSeconds < 100 & dataC$durationSeconds > 0)
us_sightings3D  <- data.frame(us_sightings3D )

centerLat = 39.82
centerLong = 98.57


# us_sightings3D$distanceFromCenter = distm (c(centerLong, centerLat), c(us_sightings3D$longitude, us_sightings3D$latitude), fun = distHaversine)

us_sightings3D <- us_sightings3D %>% rowwise() %>% 
    mutate(
      distanceFromCenter = distHaversine(c(98.57, 39.82), c(as.double(longitude),as.double(latitude))))

us_sightings3D$year = str_sub(us_sightings3D$datetime,-10,-6)

g <- plot_ly(us_sightings3D, x = ~year, y = ~durationSeconds, z = ~distanceFromCenter) %>%
  add_markers(color = ~distanceFromCenter)
g <- g %>% layout(title = "Year Found vs Distance from Center of US vs Duration")

```


```{r, echo = FALSE}
g

```

## Data Wrangling / Fixing for the 3D Scatter Plot
- I used the library(geosphere) to calculate the distance between 2 lat/long points, and used that as a column for my plot!
```{r, echo = TRUE, eval = FALSE}
us_sightings3D <- us_sightings3D %>% 
  # every row
  rowwise() %>% mutate(
      distanceFromCenter = distHaversine(c(98.57, 39.82),
          c(as.double(longitude),as.double(latitude))))
```
- Also, I used a string function to grab the data out of the longer original date format in the .csv
```{r, echo = TRUE, eval = FALSE}
us_sightings3D$year = str_sub(us_sightings3D$datetime,-10,-6)
```

## Aliens in the 90s
```{r, include = FALSE}
dataC$year = as.double(str_sub(dataC$datetime,-10,-6))
# make a copy of the us sightings

us_sightingsMap  <- filter(dataC, dataC$year >= 1990 & dataC$year <= 2000, dataC$country == 'us')


g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("gray40"),
  subunitcolor = toRGB("gray85"),
  countrycolor = toRGB("gray85"),
  countrywidth = 0.5,
  subunitwidth = 0.5
)

fig <- plot_geo(us_sightingsMap, lat = ~latitude, lon = ~longitude, color = ~state)

fig <- fig %>% add_markers(
    text = ~paste(city, state, paste("Seconds:", durationSeconds),  paste("Shape:", shape), sep = "<br />"),
    color = ~state, symbol = I("square"), size = I(8), hoverinfo = "text"
  )

fig <- fig %>% layout(
    title = 'Alien Sightings as a Map, 1990 to 2000', geo = g
  )

```


```{r, echo = FALSE, warning = FALSE}

fig

```

## Sightings over time in 3 different states

```{r, include = FALSE}
sightings  <- filter(dataC, dataC$state == 'ca' | dataC$state == 'wa' | dataC$state == 'fl', dataC$country == 'us')

sightings$year = as.double(str_sub(sightings$datetime,-10,-6))

sightings <- sightings %>% group_by(state, year) %>% mutate(currentSightings = n())

# us_sightings$year = str_sub(us_sightings3D$datetime,-10,-6)

sightings <- ungroup(sightings)

fig <- plot_ly(sightings, x = ~year, y = ~currentSightings, type = 'scatter', mode = 'markers',
  text = ~paste("Year: ", year, '<br>number of sightings:', currentSightings)) %>%
  add_lines(color = ~state)

fig <- fig %>%
  layout(
    title = "Sightings over time in 3 States",
    xaxis = list(title = "Year"),
    yaxis = list (title = "Amount of Sightings")
  )
```

```{r, echo = FALSE, warning = FALSE}
fig
```


## Data Wrangling / Fixing for the Line Graph
- Again, adding in a new column with a function is so convenient!<br>
This code makes sightings of the current year, It was able to do this because the line groups by year and then state
```{r, echo = TRUE, eval = FALSE}
sightings <- sightings %>% group_by(state, year)
%>% mutate(currentSightings = n())
```
- An interesting problem was that my lines were not showing, I found that the issue was that they were grouped,<br>
So I un-grouped before plotting
```{r, echo = TRUE, eval = FALSE}
sightings <- ungroup(sightings)
```


## Stats on Sighting Over Time (t test)

1 Population T test ( California, Washington, Florida vs. All states )
```{r, include = FALSE}
years <- sightings$year
sightingsTest <- sightings$currentSightings
```

```{r, echo = FALSE, warning = FALSE}
t.test(sightings$currentSightings, mu = mean(us_sightings$currentSightings))

```
- This would mean to reject the null Hypothesis
- One could conclude that since the p-value is near zero, that these 3 states have far more sightings per year than any other states in the USA

## Conclusion / Final Remarks

- Florida, California, and Washington are Alien spotting hot-spots
- I am sure that this data or other alien spotting data will be reviewed along with the de-classification of more Alien documents
- As technology advanced, more people came forward to report alien sightings
- Social media/smart phones played a part in the increase of sightings